Place of death in NRS COVID-19 deaths: data processing for analysis, 2024 data

### 2024-05-08 covid_nrs_place_of_death_data_2024.Rmd
just a note for myself to update script for 2024 data
modified these to include finalised 2022 data as historical, and to download the latest 2024 dataset; I've deleted & commented out the bits for the disaggregated age & sex data (currently only the split by sex/age or PoD, but not both, is available)

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

@@

## Sex & age

Note: the last sex & age split of the data provided by NRS was Week 9 of 2023, though the data can still be split using the 'overall' dataset from Open Data Scotland above!

```{r}
# weekly_deaths_sex_age <-
#   weekly_deaths_overall %>%
#   filter(age!="all" & place_of_death != "All")

# TODO: delete the below line, used for testing
# files_weekly_deaths$main_report <- "./downloaded_data/2023-03-22_weekly-deaths-23.xlsx"

sheet_covid_deaths_sex_and_age_data <- "1"
sheet_all_cause_deaths_sex_and_age_data <- "4"

# TODO: the data have changed again, now the most recent file no longer includes 2022 and 2023, but only 2023, so the cell ranges below are inaccurate and result in errors!
# 
# cell_ranges_for_sex_and_age_data <- list(
#   "All" = "A6:K110",
#   "Female" = "A113:K217",
#   "Male" = "A220:K324"
# )
cell_ranges_for_sex_and_age_data <- list(
  "All" = "A6:K58",
  "Female" = "A61:K113",
  "Male" = "A116:K168"
)

# TODO: this still doesn't split data by PoD...
weekly_deaths_sex_age <-
  full_join(
    x =
      map_dfr(
        .x = cell_ranges_for_sex_and_age_data,
        .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_covid_deaths_sex_and_age_data, range = .x), 
        .id = "sex"
      ) %>%
      pivot_longer(
        cols = `All ages`:`85+`,
        names_to = "age",
        values_to = "deaths_covid_related"
      ) %>%
      janitor::clean_names(),
    y =
      map_dfr(
        .x = cell_ranges_for_sex_and_age_data,
        .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_all_cause_deaths_sex_and_age_data, range = .x), 
        .id = "sex"
      ) %>%
      pivot_longer(
        cols = `All ages`:`85+`,
        names_to = "age",
        values_to = "deaths_all_causes"
      ) %>%
      janitor::clean_names(),
    by = c("sex", "registration_year", "week_number", "week_beginning", "age")
  ) %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  mutate(
    year = as.integer(year),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
    deaths_non_covid = deaths_all_causes - deaths_covid_related  # compute non-covid from all-cause minus covid
  ) %>%
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()

# TODO: this is obsolete now, remove!
weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  parse_year_and_week_number_from_ad_hoc_data %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(  # complete implicit missing entries and make them 0
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
```

### Sex & age separate splits

```{r}
## compute weekly data for sex and age separately
## sum over sex or age, respectively
weekly_deaths_sex <-
  weekly_deaths_sex_age %>%
  filter(age!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")

weekly_deaths_age <-
  weekly_deaths_sex_age %>%
  filter(sex!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
```

@@

## Weekly Covid as contributory vs underlying cause of death

```{r}
# TODO: need to compute that from another source for the time being
weekly_covid_contributory_vs_underlying_cause <-
  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  parse_year_and_week_number_from_ad_hoc_data %>%
  filter(
    cause_of_death %in% c("COVID-19 contributory factor", "COVID-19 underlying cause")
  ) %>%
  tidyr::complete(  # complete implicit missing entries and make them 0
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  clean_names() %>%
  filter(age!="all") %>%
  filter(sex!="all") %>%
  filter(place_of_death != "All") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, place_of_death) %>%  # sum over age & sex
  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
  ({function(x)
    bind_rows(
      x,
      x %>% 
        group_by(week_number, week_number_run_over, year, date_w_c) %>%  # sum across place of death
        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
        mutate(place_of_death = "All")
    )
  })

# TODO: this is currently commented out as it doesn't compute correctly
## check that the figures actually make sense! i.e. the two figures add up to the total covid-related figure

# stopifnot(
#   (weekly_deaths_overall %>%
#   filter(sex=="all" & age=="all") %>%
#   inner_join(weekly_covid_contributory_vs_underlying_cause) %>%
#   filter(deaths_covid_related != covid_19_contributory_factor + covid_19_underlying_cause) %>%
#   nrow) == 0
# )

annual_covid_contributory_vs_underlying_cause <-
  weekly_covid_contributory_vs_underlying_cause %>%
  group_by(year, place_of_death) %>%
  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
  ({function(x)
    bind_rows(
      x %>% mutate(year = as.character(year)),
      x %>% 
        group_by(place_of_death) %>%  # sum across eyars
        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
        mutate(year = "2020-2021")
    )
  }) %>%
  mutate(deaths_covid_related = covid_19_contributory_factor + covid_19_underlying_cause) %>%
  mutate(
    proportion_contributory = covid_19_contributory_factor / deaths_covid_related,
    proportion_underlying = covid_19_underlying_cause / deaths_covid_related
    )
```

@@

```{r}
sheet_cause_of_death_data <- "8"

@@

## this was for only 2022 data
## ...and is correct again for 2023
cell_ranges_for_causes_of_death_data <- list(
  "All" = "A6:V58",
  "Care home" = "A61:V113",
  "Home & other non-institution" = "A116:V168",
  "Hospital" = "A171:V223",
  "Other" = "A226:V278"
)

# TODO: this will eventually only have 2023 data, change when that happens
weekly_deaths_cause_of_death_2022_onwards <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$")) %>%  # remove excess deaths, not currently used in analysis
  # select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "number_of_deaths"),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  pivot_wider(names_from = measure_type, values_from = number_of_deaths)
  # filter(!is.na(number_of_deaths))   # this will also remove the historic mean values, so I've opted not to remove blank entries!
```

### Data from 2020 & 2021

@@

```{r}
# TODO: deal with 2022 data here, now that it's 'historical'; this will involve loading the last released data for 2022 released in the weekly releases for 2023, I assume somewhere mid-2023, like it was for 2021 data in 2022 releases
cell_ranges_for_causes_of_death_data_2022 <- list(
  "All" = "A6:V58",
  "Care home" = "A61:V113",
  "Home & other non-institution" = "A116:V168",
  "Hospital" = "A171:V223",
  "Other" = "A226:V278"
)

```

@@

historical_weekly_average_la <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_la, 
        current_data = weekly_deaths_la,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, ref_area
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates

historical_weekly_average_sex_age <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_sex_age, 
        current_data = weekly_deaths_sex_age,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, age, sex
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates

historical_weekly_average_age <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_age, 
        current_data = weekly_deaths_age,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, age
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates

historical_weekly_average_sex <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_sex, 
        current_data = weekly_deaths_sex,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, sex
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
```

@@

historical_annual_average_sex_age <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_sex_age, 
        current_data = weekly_deaths_sex_age, 
        year_range = years_in_historic_period
      ) %>%
        mutate(historic_period = historic_period)
    }
  ) %>%
  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
  compute_historical_annual_average_deaths(sex, age)

historical_annual_average_sex <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_sex, 
        current_data = weekly_deaths_sex, 
        year_range = years_in_historic_period
      ) %>%
        mutate(historic_period = historic_period)
    }
  ) %>%
  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
  compute_historical_annual_average_deaths(sex)

historical_annual_average_age <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_age, 
        current_data = weekly_deaths_age, 
        year_range = years_in_historic_period
      ) %>%
        mutate(historic_period = historic_period)
    }
  ) %>%
  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
  compute_historical_annual_average_deaths(age)
```

@@

## here I separately merge the datasets for cause of death data from different years
## I treat covid deaths separately to recalculate a realistic esstimate of average deaths in prior years
weekly_deaths_cause_of_death_noncovid <-
  bind_rows(
    weekly_deaths_cause_of_death_2020,
    weekly_deaths_cause_of_death_2021,
    weekly_deaths_cause_of_death_2022_onwards
  ) %>%
  filter(cause_of_death != "COVID-19")

weekly_deaths_cause_of_death_covid_before_processing <-
  bind_rows(
    weekly_deaths_cause_of_death_2020,
    weekly_deaths_cause_of_death_2021,
    weekly_deaths_cause_of_death_2022_onwards
  )

@@

merged_deaths_la <-
  weekly_deaths_la %>%  # use data with weekly movingincluded
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_la, by = c("week_number","ref_area","place_of_death", "historic_period"))

merged_deaths_sex_age <-
  weekly_deaths_sex_age %>%  # use data with weekly movingincluded
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_sex_age, by = c("place_of_death", "week_number", "sex", "age", "historic_period"))

merged_deaths_sex <-
  weekly_deaths_sex %>%
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_sex, by = c("place_of_death", "week_number", "sex", "historic_period"))

merged_deaths_age <-
  weekly_deaths_age %>%
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_age, by = c("place_of_death", "week_number", "age", "historic_period"))

@@

total_deaths_sex <-
  merged_deaths_sex %>%
  compute_total_deaths_in_year(sex) %>%
  left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
  compute_ratio_of_annual_deaths_to_historical %>%
  remove_invalid_historic_period_for_year()

total_deaths_age <-
  merged_deaths_age %>%
  compute_total_deaths_in_year(age) %>%
  left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
  compute_ratio_of_annual_deaths_to_historical %>%
  remove_invalid_historic_period_for_year()
```

@@

weeks_available_2024_disaggregated <- max(merged_deaths_age$week_number[merged_deaths_age$year==2024])
most_recent_date_available_2024_disaggregated <- max(merged_deaths_age$date_w_c)
```

@@

merged_deaths_la <-
  compute_moving_average_for_current_data(
    current_data = merged_deaths_la, 
    historical_data = 
      weekly_deaths_2015_2019_la, 
    num_weeks_to_average = 4)

merged_deaths_sex <-
  compute_moving_average_for_current_data(
    current_data = merged_deaths_sex, 
    historical_data = 
      weekly_deaths_2015_2019_sex, 
    num_weeks_to_average = 4)

merged_deaths_age <-
  compute_moving_average_for_current_data(
    current_data = merged_deaths_age, 
    historical_data = 
      weekly_deaths_2015_2019_age, 
    num_weeks_to_average = 4)

merged_deaths_sex_age <-
  compute_moving_average_for_current_data(
    current_data = merged_deaths_sex_age, 
    historical_data = 
      weekly_deaths_2015_2019_sex_age, 
    num_weeks_to_average = 4)
```

@@

# TODO: other thigns like excess death & historical rates by geography

```{r}
# TODO: copy some of the old calculations over
```

>>>>

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
### remotes::install_github("datasciencescotland/opendatascot", force = TRUE)  # this is the old repo
# devtools::install_github("jsphdms/opendatascot")
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

@@

@@

@@

```{r}
# TODO: delete the below line, used for testing
files_weekly_deaths$main_report <- "../../../Downloads/weekly-deaths-24-week-17.xlsx"

sheet_cause_of_death_data <- "8"

## updated for 2024, this has all of 2023 plus incomplete 2024 data, with blanks up to week 52
## note that there are fewer columns now because excess deaths have been moved to a separate spreadsheet
# TODO: 2024 has 52 ISOweeks, but 2024-12-30 is W01 of 2025, or the 53rd week of 2024, so this may change
cell_ranges_for_causes_of_death_data <- list(
  "All" = "A6:J110",
  "Care home" = "A113:J217",
  "Home & other non-institution" = "A220:J324",
  "Hospital" = "A327:J431",
  "Other" = "A434:J538"
)

@@

## this was for only 2022 data
## ...and is correct again for 2023
# cell_ranges_for_causes_of_death_data <- list(
#   "All" = "A6:V58",
#   "Care home" = "A61:V113",
#   "Home & other non-institution" = "A116:V168",
#   "Hospital" = "A171:V223",
#   "Other" = "A226:V278"
# )

weekly_deaths_cause_of_death_2023_onwards <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  janitor::remove_empty(which = "cols") %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$")) %>%  # remove excess deaths, not currently used in analysis
  # select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "number_of_deaths"),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  pivot_wider(names_from = measure_type, values_from = number_of_deaths)
  # filter(!is.na(number_of_deaths))   # this will also remove the historic mean values, so I've opted not to remove blank entries!
```

### Data from previous years, 2020-2022

@@

```{r}
# TODO: deal with 2022 data here, now that it's 'historical', I load the last known 2023 file that included 2022 data, 2023-07-19_weekly-deaths-23.xlsx
cell_ranges_for_causes_of_death_data_2022 <- list(
  "All" = "A6:V58",
  "Care home" = "A61:V113",
  "Home & other non-institution" = "A116:V168",
  "Hospital" = "A171:V223",
  "Other" = "A226:V278"
)

weekly_deaths_cause_of_death_2022 <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data_2022,
    .f = ~readxl::read_excel(files_historical_deaths$`2022`, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$")) %>%  # remove excess deaths, not currently used in analysis
  # select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "number_of_deaths"),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  pivot_wider(names_from = measure_type, values_from = number_of_deaths) %>%
  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
```

@@

historical_weekly_average_la <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_la, 
        current_data = weekly_deaths_la,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, ref_area
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
# 
# historical_weekly_average_sex_age <-
#   pmap_dfr(
#     .l = historic_period_by_year,
#     .f = function(year, historic_period, years_in_historic_period) {
#       get_historic_and_current_data_from_year_range(
#         historic_data = weekly_deaths_2015_2019_sex_age,
#         current_data = weekly_deaths_sex_age,
#         year_range = years_in_historic_period
#       ) %>%
#       create_week_53_data_for_years_with_52_weeks() %>%
#       group_by(
#         week_number, place_of_death, age, sex
#       ) %>%
#       deaths_mean_min_max %>%
#       mutate(historic_period = historic_period)
#     }
#   ) %>%
#   distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
# 
# historical_weekly_average_age <-
#   pmap_dfr(
#     .l = historic_period_by_year,
#     .f = function(year, historic_period, years_in_historic_period) {
#       get_historic_and_current_data_from_year_range(
#         historic_data = weekly_deaths_2015_2019_age, 
#         current_data = weekly_deaths_age,
#         year_range = years_in_historic_period
#       ) %>%
#       create_week_53_data_for_years_with_52_weeks() %>%
#       group_by(
#         week_number, place_of_death, age
#       ) %>%
#       deaths_mean_min_max %>%
#       mutate(historic_period = historic_period)
#     }
#   ) %>% 
#   distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
# 
# historical_weekly_average_sex <-
#   pmap_dfr(
#     .l = historic_period_by_year,
#     .f = function(year, historic_period, years_in_historic_period) {
#       get_historic_and_current_data_from_year_range(
#         historic_data = weekly_deaths_2015_2019_sex, 
#         current_data = weekly_deaths_sex,
#         year_range = years_in_historic_period
#       ) %>%
#       create_week_53_data_for_years_with_52_weeks() %>%
#       group_by(
#         week_number, place_of_death, sex
#       ) %>%
#       deaths_mean_min_max %>%
#       mutate(historic_period = historic_period)
#     }
#   ) %>% 
#   distinct  # the 2020 and 2021 data get identical historical periodfs & values, so I remove duplicates
```

@@

# historical_annual_average_sex_age <-
#   pmap_dfr(
#     .l = historic_period_by_year,
#     .f = function(year, historic_period, years_in_historic_period) {
#       get_historic_and_current_data_from_year_range(
#         historic_data = weekly_deaths_2015_2019_sex_age, 
#         current_data = weekly_deaths_sex_age, 
#         year_range = years_in_historic_period
#       ) %>%
#         mutate(historic_period = historic_period)
#     }
#   ) %>%
#   distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
#   compute_historical_annual_average_deaths(sex, age)
# 
# historical_annual_average_sex <-
#   pmap_dfr(
#     .l = historic_period_by_year,
#     .f = function(year, historic_period, years_in_historic_period) {
#       get_historic_and_current_data_from_year_range(
#         historic_data = weekly_deaths_2015_2019_sex, 
#         current_data = weekly_deaths_sex, 
#         year_range = years_in_historic_period
#       ) %>%
#         mutate(historic_period = historic_period)
#     }
#   ) %>%
#   distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
#   compute_historical_annual_average_deaths(sex)
# 
# historical_annual_average_age <-
#   pmap_dfr(
#     .l = historic_period_by_year,
#     .f = function(year, historic_period, years_in_historic_period) {
#       get_historic_and_current_data_from_year_range(
#         historic_data = weekly_deaths_2015_2019_age, 
#         current_data = weekly_deaths_age, 
#         year_range = years_in_historic_period
#       ) %>%
#         mutate(historic_period = historic_period)
#     }
#   ) %>%
#   distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
#   compute_historical_annual_average_deaths(age)
```

@@

## here I separately merge the datasets for cause of death data from different years
## I treat covid deaths separately to recalculate a realistic estimate of average deaths in prior years
weekly_deaths_cause_of_death_noncovid <-
  bind_rows(
    weekly_deaths_cause_of_death_2020,
    weekly_deaths_cause_of_death_2021,
    weekly_deaths_cause_of_death_2022,
    weekly_deaths_cause_of_death_2023_onwards
  ) %>%
  filter(cause_of_death != "COVID-19")

weekly_deaths_cause_of_death_covid_before_processing <-
  bind_rows(
    weekly_deaths_cause_of_death_2020,
    weekly_deaths_cause_of_death_2021,
    weekly_deaths_cause_of_death_2022,
    weekly_deaths_cause_of_death_2023_onwards
  )

@@

merged_deaths_la <-
  weekly_deaths_la %>%  # use data with weekly movingincluded
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_la, by = c("week_number","ref_area","place_of_death", "historic_period"))
# 
# merged_deaths_sex_age <-
#   weekly_deaths_sex_age %>%  # use data with weekly movingincluded
#   mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
#   inner_join(historical_weekly_average_sex_age, by = c("place_of_death", "week_number", "sex", "age", "historic_period"))
# 
# merged_deaths_sex <-
#   weekly_deaths_sex %>%
#   mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
#   inner_join(historical_weekly_average_sex, by = c("place_of_death", "week_number", "sex", "historic_period"))
# 
# merged_deaths_age <-
#   weekly_deaths_age %>%
#   mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
#   inner_join(historical_weekly_average_age, by = c("place_of_death", "week_number", "age", "historic_period"))

@@

# total_deaths_sex <-
#   merged_deaths_sex %>%
#   compute_total_deaths_in_year(sex) %>%
#   left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
#   compute_ratio_of_annual_deaths_to_historical %>%
#   remove_invalid_historic_period_for_year()
# 
# total_deaths_age <-
#   merged_deaths_age %>%
#   compute_total_deaths_in_year(age) %>%
#   left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
#   compute_ratio_of_annual_deaths_to_historical %>%
#   remove_invalid_historic_period_for_year()
```

@@

weeks_available_2024_disaggregated <- max(merged_deaths_cause_of_death$week_number[merged_deaths_cause_of_death$year==2024])
most_recent_date_available_2024_disaggregated <- max(merged_deaths_cause_of_death$date_w_c)
```

@@

merged_deaths_la <-
  compute_moving_average_for_current_data(
    current_data = merged_deaths_la, 
    historical_data = 
      weekly_deaths_2015_2019_la, 
    num_weeks_to_average = 4)
# 
# merged_deaths_sex <-
#   compute_moving_average_for_current_data(
#     current_data = merged_deaths_sex, 
#     historical_data = 
#       weekly_deaths_2015_2019_sex, 
#     num_weeks_to_average = 4)

# merged_deaths_age <-
#   compute_moving_average_for_current_data(
#     current_data = merged_deaths_age, 
#     historical_data = 
#       weekly_deaths_2015_2019_age, 
#     num_weeks_to_average = 4)

# merged_deaths_sex_age <-
#   compute_moving_average_for_current_data(
#     current_data = merged_deaths_sex_age, 
#     historical_data = 
#       weekly_deaths_2015_2019_sex_age, 
#     num_weeks_to_average = 4)
```

@@
